Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-11-06 paper/manuscript.tex
Minor tweaks from @sboresch and @ppxasjsm

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[most]{tcolorbox}
\usepackage{enumitem,amssymb}
\usepackage{textgreek}
\usepackage{changepage}
\usepackage{verbatim}

@@

Next, we outline the application of the chemical potential statistical mechanics formalism for modeling key biophysical processes: solvation, solvent partitioning, and non-covalent molecular binding. Generally, turning statistical mechanics identities into computable quantities by molecular simulations involves expressing ratios of partition functions in terms of ensemble averages. In the present context, the quantity being averaged often turns out to be the exponential of a perturbation energy function (the \textit{Zwanzig relationship}~\cite{zwanzig1954hightemperature}), which is efficiently evaluated using the stratification algorithms and free energy estimators discussed in Sections \ref{subsec:formulations-abfe-rbfe}, \ref{subsec:numerical-implementations},  and \ref{subsec:estimators} rather than directly. Here we assume that an implementation is well-formed once it is expressed as a combination of exponential averages, assuming they can be evaluated in practice. As in the previous section, we use the solvent potential of mean force to lighten the notation when possible. The solvent degrees of freedom can be restored  in the resulting expressions using Eq.~(\ref{eq:solvent-pmf-def}).

@@

The standard molar Gibbs free energy of the solvation of a gas species $A$ in a solvent $v$
\begin{equation}
A(\mathrm{g})  \leftrightharpoons A(v)
\end{equation}
is
\begin{equation}
  \Delta G^\circ_{\mathrm{slv}} = \mu^\circ_{A(v)}  - \mu^\circ_{A(g)} = -k_B T \ln \frac{P^\circ}{C^\circ k_B T} - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}}
  \label{eq:DG0-solvation-def}
\end{equation}
where we used Eq.~(\ref{eq:chemical-potential-standard}) for the solution and  Eq.~(\ref{eq:chemical-potential-definition}) for the gas (assumed ideal) at the standard gas concentration $C = P^\circ/k_B T$ where $P^\circ = 1$ bar is the standard pressure, and Eq.~(\ref{eq:intra-qpu-def}) noting that all momentum terms cancel. The first term in Eq.~(\ref{eq:DG0-solvation-def}) is an ideal term $\Delta G^\circ_{\mathrm{slv,ideal}}$ that accounts for the difference of standard states for solution and gases. The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations (see below). Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.\cite{thompson2004new}

However, equilibrium experimental gas solvation data is most often available in terms of Henry's constants $k_H$. Setting $\Delta G_{\mathrm{slv}} = 0$ when the gas partial pressure and the solute's concentrations are equal to their equilibrium values $P_A$ and $C_A$, respectively, we obtain
\begin{equation}
  P_A = k_H C_A
  \label{eq:Henrys-law}
\end{equation}
where
\begin{equation}
  k_H = k_B T \frac{z_{A(\mathrm{g})}}{z_{A(v)}} \, .
  \label{eq:Henrys-constant}
\end{equation}
The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{cabani1981group,nicholls2008predicting,mobley2014freesolv}
\begin{equation}
\Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}} = k_B T \ln \frac{k_H}{k_B T} \, .
\label{eq:DG-solvation-excess}
\end{equation}
Because it involves only intramolecular partition functions, Eq.~(\ref{eq:DG-solvation-excess}) states that the excess free energy of solvation corresponds to the process of moving the solute from a fixed position and orientation in the gas to a fixed position and orientation in the solvent. This is sometimes referred to as the solvation free energy in the Ben-Naim standard state.~\cite{bennaim1984solvation}

@@

Combining Eqs.~(\ref{eq:DG-solvation-excess}) and (\ref{eq:zratio-solv-average}) yields\cite{widom1982potential} 
\begin{equation}
  \Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \langle e^{-\beta \Delta U} \rangle_0 = k_B T \ln \langle e^{\beta \Delta U} \rangle_1
  \label{eq:zratio-solv-dgexc}
\end{equation}
The objective of alchemical solvation free energy calculations is to implement Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols and alchemical intermediate states as  discussed in section \ref{subsec:numerical-implementations}.

@@

Eq.~(\ref{eq:DG0-binding-constant-def}) is the basis of a variety of alchemical computational implementations that have have been developed to study molecular binding by alchemical means. Some of these are discussed in Section \ref{subsec:formulations-abfe-rbfe}. Generally, to formulate the statistical mechanics expression of the binding constant Eq.~(\ref{eq:DG0-binding-constant-def}) as an ensemble average, we first need to ensure that the configurational integrals in the numerator and denominator are over the same set of degrees of freedom. Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand. To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
\begin{equation} \label{eq:Vsite-def}
    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
\end{equation}
which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent. The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart. The result is
\begin{equation}\label{eq:DG0-binding-constant-def-average}
  K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0
\end{equation}
where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
\begin{equation}\label{eq:pert-energy-def}
  \Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
\end{equation}
is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble. Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
\end{equation}
which, expressed in free energy units, yields
\begin{eqnarray}\label{eq:DG0-def-Vsite}
  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
\end{eqnarray}

The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, the binding free energy that would be observed in the absence of molecular interactions. Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component that depends on the specific molecular composition of the system and on the corresponding interactions. The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.

Eq~(\ref{eq:DG0-binding-constant-def-average-noOmega}) instructs to collect samples of the solvent-averaged binding energy $\Delta \Psi$ in the ensemble where ligand and receptor are not interacting. This can be done directly for simple solutes when the solvent is modeled implicitly.\cite{gallicchio2014virtual} However, with explicit solvation specialized techniques are required. We will discuss two: Double-Decoupling and Alchemical Transfer in section \ref{subsec:formulations-abfe-rbfe}.

@@

\subsection{Alchemical Binding Free Energy Methods Formulations}
\label{subsec:formulations-abfe-rbfe}

\subsubsection{Absolute binding free energy (ABFE)}

In the alchemical literature, the term ``absolute binding free energy'' is another name for the standard binding free energy $\Delta G^\circ_b = - k_B T \ln K_b$. Here, we discuss two methods, Double-Decoupling (DDM) and Alchemical Transfer (ATM), used to evaluate it by alchemical computer simulations.

\paragraph{Double Decoupling}

The double-decoupling formula is derived from the statistical mechanics theory
by inserting the definition of the solvent potential of mean force [Eq.~(\ref{eq:solvent-pmf-def})] in each of the configurational partition functions in  Eq.~(\ref{eq:DG0-binding-constant-def}). For example, after expanding the solvent potential of mean force, the configurational partition function of the complex becomes
\begin{equation}\label{eq:partfunct-ratio-explsolv-1}
z_{RL(v)} = \frac{Z_{N_v, RL}}{Z_{N_v}}
\end{equation}
where
\begin{equation}\label{eq:partfunct-ratio-explsolv-2}
Z_{N_v, RL} = \int d\vec{r}_v dx_R dx_L d\zeta I(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
\end{equation}
is the configurational partition function of the complex in a fixed position and orientation in $N_v$ solvent molecules, and $Z_{N_v}$ is the configurational partition function of the pure solvent. This process leads to an expression for the binding constant in terms of products of ratios of partition functions analogous to those encountered in Section \ref{sec:theory-solvation} to model solvation
\begin{equation}
  K_b = \frac{C^\circ}{8 \pi^2} \frac{Z_{N_v,RL}}{Z_{N_v,R} z_{L(g)}} \frac{Z_{N_v} z_{L(g)}}{Z_{N_v,L}}
  \label{eq:ddm-ratios-1}
\end{equation}
where we multiplied and divided by the internal configurational partition function of the ligand in vacuum
\begin{equation}
z_{L(g)} = \int dx_L e^{-\beta U(x_L)}
\end{equation}
We recognize that the second ratio of partition functions in Eq.~(\ref{eq:ddm-ratios-1}) is the inverse of Eq.~(\ref{eq:zratio-solv-average}), hence it is related to the free energy $\Delta G_1$ of decoupling the ligand from the solvent
\begin{equation}
\frac{Z_{N_v} z_{L(g)}}{Z_{N_v,L}} = \langle e^{\beta \Delta U}  \rangle_{L(v)} = e^{-\beta \Delta G_1}
\end{equation}
where $\Delta U$ is the coupling energy between the ligand and the solvent and $\langle \ldots \rangle_{L(v)}$ is an average over the ligand solution ensemble [see Eq.~(\ref{eq:zratio-solv-dgexc})].
The ratio of partition functions in Eq.~(\ref{eq:ddm-ratios-1}) involving the ligand-receptor complex is expressed as an ensemble average by multiplying it and dividing it by Eq.~(\ref{eq:Vsite-def}) in the same way that Eq.~(\ref{eq:DG0-binding-constant-def}) is cast as Eq.~(\ref{eq:DG0-binding-constant-def-average}). The result is the ratio of the partition functions of the complex in solvent, $Z_{N_v,RL}$, and the partition function $Z_{N_v,R + L(g)}$  of the same system when the ligand is decoupled from the
receptor and the solvent but is restrained into the binding site region ($I(\zeta) = 1$)
\begin{equation}
\frac{Z_{N_v,RL}}{Z_{N_v,R+L} z_{L(g)}} = \frac{8 \pi^2 V_{\rm site} Z_{N_v,RL}}{Z_{N_v,R+L(g)}} = 8 \pi^2 V_{\rm site} \langle e^{\beta \Delta U}  \rangle_{RL(v)} = e^{\beta \Delta G_2}
\end{equation}
Collecting the terms, the double-decoupling standard binding free energy is finally expressed as
\begin{equation}\label{eq:ddm-dg1-dg2}
\Delta G^\circ_b = -k_B T \ln K_b = -k_B T \ln C^\circ V_{\rm site} + (\Delta G_1 - \Delta G_2)
\end{equation}
which includes the ideal term [see Eq.~(\ref{eq:DG0-def-Vsite})], and the difference of decoupling free energies of the ligand from the solvent and from the binding site of the receptor in solution. As illustrated in Figure \ref{fig:fig_absolute_thermodynamic_cycle}, the decoupling free energies are commonly computed in separate simulations using the alchemical techniques discussed in the following sections.

\paragraph{Alchemical Transfer}

In the alchemical transfer approach, the standard binding free energy is obtained by a coordinate translation that brings the ligand from the solvent to the receptor binding site~\cite{wu2021alchemical} rather than by decoupling it from the system, as in the double-decoupling method. The method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) by inserting the solvent degrees of freedom similarly to the double-decoupling method, except that the product $z_{R(v)} z_{L(v)}$ of partition functions of the unbound state is represented collectively as one explicit solvent system in which the ligand is placed in a region spanned by an indicator function $I^\ast(\zeta)$ identical to the one used to define the binding site region but placed at some arbitrarily large distance from the receptor:
\begin{equation}
  z_{R(v)} z_{L(v)} =
  \frac{
    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }{
    8 \pi^2 V_{\rm site} Z_{N_v}
  }
\end{equation}
where, as above, we multiplied and divided by the volume of the binding site, assuming for simplicity that the indicator function does not depend on the orientational degrees of freedom of the ligand. Combining this result with Eq.~(\ref{eq:partfunct-ratio-explsolv-1}) for the complex and canceling the common partition function of the pure solvent, leads with Eq.~(\ref{eq:DG0-binding-constant-def})  to an expression for the binding constant in terms of ratios of partition functions in explicit solvent that differ only by the location of the indicator function (in the binding site or in the solvent region):
\begin{equation}\label{eq:atm-partfunc-ratio}
  K_b = C^\circ V_{\rm site} \frac{
    \int d\vec{r}_v dx_R dx_L d\zeta I(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }
\end{equation}
The idea of alchemical transfer, illustrated in Figure \ref{fig:atm-illustration}A, is to cast Eq.~(\ref{eq:atm-partfunc-ratio}) as an ansemble average by a change of coordinates noting that $I(\zeta) = I^\ast(\zeta - \vec{h})$ where $\vec{h}$ is the fixed vector distance between the binding site and solvent locations spanned by the two indicator functions. The symbol $\zeta$ represents the position and orientation of the ligand relative to the receptor, so here $\zeta - \vec{h}$ is the operation of shifting the position of the ligand by the displacement $\vec{h}$ without changing its orientation. Hence, by applying the change of variable $\zeta \rightarrow \zeta - \vec{h}$ to the integral in the numerator of Eq.~(\ref{eq:atm-partfunc-ratio}), we arrive at the following expression for the binding constant, equivalent to Eq.~(\ref{eq:DG0-binding-constant-def-average-noOmega}):
\begin{equation}\label{eq:atm-average-abfe}
K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta U} \rangle_0
\end{equation}
where
\begin{equation}\label{eq:atm-deltaU-def}
\Delta U(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U(x_R, x_L, \zeta, \vec{r}_v)
\end{equation}
is the perturbation energy of the method, and the ensemble average $\langle \ldots \rangle_0$ is over the unbound ensemble when the ligand is limited to the solvent region far away from the receptor. Eq.~(\ref{eq:atm-average-abfe}) instructs to simulate the unbounded system where the ligand is limited to a region in the solvent and to sample the change of potential energy $\Delta U$ resulting from transferring the ligand from the solvent to the receptor binding site by a fixed displacement of the ligand as a whole.

While Eq.~(\ref{eq:atm-average-abfe}) is formally rigorous, as discussed in Section \ref{subsec:numerical-implementations} it is numerically necessary to divide the alchemical transfer ABFE calculation in explicit solvent into two legs (Figure \ref{fig:atm_alchemical_intermediate})~\cite{wu2021alchemical,azimi2022relative} that connect the unbound and bound states by a suitable alchemical intermediate state (Figure \ref{fig:atm_alchemical_intermediate}). Current implementations use an alchemical intermediate state midway between the bound and unbound states with the symmetric hybrid potential energy function
\begin{equation}\label{eq:atm-intermediate-pot}
U_{1/2}(x) = \frac12 [ U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) + U(x_R, x_L, \zeta, \vec{r}_v)  ] 
\end{equation}
To include the alchemical intermediate, Eq.~(\ref{eq:atm-partfunc-ratio}) is multiplied and divided by the partition function of the system with the hybrid potential energy function in Eq.~(\ref{eq:atm-intermediate-pot}), each serving as the sampling state for each leg. The expression for the binding constant becomes a ratio of ensemble averages
\begin{equation}\label{eq:atm-average-abfe-intermediate}
  K_b = C^\circ V_{\rm site} \frac{
    \langle e^{-\beta \Delta U_1} \rangle_{1/2}
  }{
    \langle e^{-\beta \Delta U_2} \rangle_{1/2}
  }
\end{equation}
where $\Delta U_1$ and $\Delta U_2$ are the perturbation energies of the bound and unbound states relative to the alchemical intermediate. For example,
\begin{equation}
\Delta U_1(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U_{1/2}(x)
\end{equation}
Setting,
\begin{equation}
\Delta G_1 = -k_B T \ln \langle e^{-\beta \Delta U_1} \rangle_{1/2}
\end{equation}
and similarly for the second leg, we recover Eq.~(\ref{eq:ddm-dg1-dg2}) obtained for double-decoupling. This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer. However, alchemical transfer simulates both legs with a single simulation system while double-decoupling accomplishes them in separate simulations.

\subsubsection{Relative binding free energy (RBFE)}

In many cases, the quantity of interest is the change in binding affinity $\Delta \Delta G_b$ between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})) for the same receptor. If the ligands are sufficiently similar, evaluating the Relative Binding Free Energy (RBFE) is often more convenient than taking the difference of their ABFEs. Formally, these approaches seek to compute the ratio $K_b(B)/K_b(A)$ of the binding constants of the two ligands rather than each binding constant individually. Here we present an RBFE formulation based on the double-decoupling ABFE approach, which we call double-transformation, and the RBFE formulation based on alchemical transfer.

\paragraph{Double Transformation for RBFE estimation}

To derive an RBFE formulation from double-decoupling, we write expressions for the binding constants of $A$ and $B$ using Eq.~(\ref{eq:ddm-ratios-1}) and taking their ratio. While doing so, the standard state, pure solvent, and gas-phase terms cancel, and we obtain
\begin{equation}\label{eq:double-transformation-Kb-rbfe}
\frac{K_b(B)}{K_b(A)} = \frac{Z_{N_v,RB}}{Z_{N_v,RA}} \frac{Z_{N_v,A}}{Z_{N_v,B}}
\end{equation}
where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution. Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor (Figure \ref{fig:fig_binding_thermodynamic_cycle}):
\begin{equation}\label{eq:double-transformation-dg-rbfe}
\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,A}}{Z_{N_v,B}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
\end{equation}
In practice, each ratio of partition functions must first be turned into an ensemble average to evaluate it. Similar to the preceding case, this step requires manipulating the integrals so that they are carried over the same set of degrees of freedom. As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies capable of describing both ligands. For example, among other features, alchemical topologies include atoms common to both ligands and dummy atoms that exist in one but not in the other ligand (Figure \ref{fig:fig_topology}). In one limit, called dual coordinate dual topology in Section \ref{sec:relative-fe-protocol}, all atoms of both ligands are present, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system. A key aspect of constructing a suitable alchemical topology is the requirement that the additional degrees of freedom do not affect the RBFE estimate.

@@

\paragraph{Alchemical Transfer for RBFE Estimation}

The RBFE alchemical transfer formulation is very similar to the ABFE formulation. The simulation system includes both ligands, one in solution and the other in the binding site, and the coordinate transformation involves displacing them in opposite directions so that they switch places. Formally,~\cite{azimi2022relative} the method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) for each binding constant
\begin{equation}
\frac{K_b(B)}{K_b(A)} = \frac{z_{RB(v)} z_{A(v)}}{z_{B(v)} z_{RA(v)}}
\end{equation}
and noting that the numerator represents a system with $B$ bound and $A$ unbound, and in the denominator, the roles are switched. Next, the solvent PMF is used to indicate the solvent degrees of freedom explicitly, and the indicator functions $I^\ast(\zeta_A)$ and $I^\ast(\zeta_b)$ are introduced to keep the unbound ligands in the solvent, similarly to Eq.~(\ref{eq:atm-partfunc-ratio}). The result is
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-1}
  \frac{K_b(B)}{K_b(A)} = \frac{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I^\ast(\zeta_A) I(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }
\end{equation}
where $x_A$ and $x_B$ are the internal coordinates of the ligands and $\zeta_A$, $\zeta_B$ are their external coordinates relative to the receptor.  Next we apply the change of variables $\zeta_B \rightarrow \zeta_B + \vec{h}$ and $\zeta_A \rightarrow \zeta_A - \vec{h}$ to the partition function in the numerator to obtain
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-2}
  \frac{K_b(B)}{K_b(A)} = \frac{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
  }
\end{equation}
which can be written as an ensemble average
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-average}
\frac{K_b(B)}{K_b(A)} = \langle e^{-\beta \Delta U}\rangle_{RA+B}
\end{equation}
where
\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-pert}
\Delta U = U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v) - U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)
\end{equation}
is the perturbation energy for displacing the two ligands in opposite directions, and the ensemble average $\langle \ldots \rangle_{RA+B}$ is carried out with $A$ bound and $B$ in the solvent. Except for the coordinate displacement, Eq.~(\ref{eq:atm-partfunc-ratio-rbfe-average}) is implemented similarly to the alchemical transfer ABFE protocol, including the use of the symmetric alchemical intermediate as in Eq.~(\ref{eq:atm-average-abfe-intermediate}).

\subsection{Numerical Implementations of Alchemical Transformations}
\label{subsec:numerical-implementations}

@@

\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[most]{tcolorbox}
\usepackage{enumitem,amssymb}
\usepackage{textgreek}
\usepackage{changepage}

@@

@@

The standard molar Gibbs free energy of the solvation of a gas species $A$ in a solvent $v$
\begin{equation}
A(\mathrm{g})  \leftrightharpoons A(v)
\end{equation}
is
\begin{equation}
  \Delta G^\circ_{\mathrm{slv}} = \mu^\circ_{A(v)}  - \mu^\circ_{A(g)} = -k_B T \ln \frac{P^\circ}{C^\circ k_B T} - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}}
  \label{eq:DG0-solvation-def}
\end{equation}
where we used Eq.~(\ref{eq:chemical-potential-standard}) for the solution and  Eq.~(\ref{eq:chemical-potential-definition}) for the gas (assumed ideal) at the standard gas concentration $C = P^\circ/k_B T$ where $P^\circ = 1$ bar is the standard pressure, and Eq.~(\ref{eq:intra-qpu-def}) noting that all momentum terms cancel. The first term in Eq.~(\ref{eq:DG0-solvation-def}) is an ideal term $\Delta G^\circ_{\mathrm{slv,ideal}}$ that accounts for the difference of standard states for solution and gases. The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations (see below). Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.

However, equilibrium experimental gas solvation data is most often available in terms of Henry's constants $k_H$. Setting $\Delta G_{\mathrm{slv}} = 0$ when the gas partial pressure and the solute's concentrations are equal to their equilibrium values $P_A$ and $C_A$, respectively, we obtain
\begin{equation}
  P_A = k_H C_A
  \label{eq:Henrys-law}
\end{equation}
where
\begin{equation}
  k_H = k_B T \frac{z_{A(\mathrm{g})}}{z_{A(v)}} \, .
  \label{eq:Henrys-constant}
\end{equation}
The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{nicholls2008predicting,mobley2014freesolv}
\begin{equation}
\Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}} = k_B T \ln \frac{k_H}{k_B T} \, .
\label{eq:DG-solvation-excess}
\end{equation}
Because it involves only intramolecular partition functions, Eq.~(\ref{eq:DG-solvation-excess}) states that the excess free energy of solvation corresponds to the process of moving the solute from a fixed position and orientation in the gas to a fixed position and orientation in the solvent. This is sometimes referred to as the solvation free energy in the Ben-Naim standard state.~\cite{bennaim1984solvation}

@@

Combining Eqs.~(\ref{eq:DG-solvation-excess}) and (\ref{eq:zratio-solv-average}) yields\cite{widom1982potential} 
\begin{equation}
  \Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \langle e^{-\beta \Delta U} \rangle_0 = k_B T \ln \langle e^{\beta \Delta U} \rangle_1
  \label{eq:zratio-solv-dgexc}
\end{equation}
The objective of alchemical solvation free energy calculations is to evaluate Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols discussed in the following sections.

@@

Eq.~(\ref{eq:DG0-binding-constant-def}) is the basis of a variety of alchemical computational implementations discussed in the following sections that have have been developed to study molecular binding by alchemical means.

@@

\subsection{Alchemical free energy calculations yield predictions that do not require direct simulation of binding/unbinding events}

In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
\begin{equation}\label{eq:delta-delta-G-physical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
\end{split}
\end{equation}
Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}.
This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
\begin{equation}\label{eq:delta-delta-G-alchemical}
\begin{split}
    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
\end{split}
\end{equation}
where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
Eq.~\ref{eq:delta-delta-G-alchemical} and Fig.~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
Note that, when $A$ and $B$ have a different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.

@@

@@

\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimized structure of a solute can be generated with a simple tool such as Open Babel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.

\paragraph{The choice of solvent force field may impact the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation of the system ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced different results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.
